---
title: 'APTS: Statistical Computing Assessment'
summary: Go through the solutions for the Statistical Computing assessment
author: André Felipe Berdusco Menezes
date: '2023-01-30'
draft: false
categories:
  - statistical computing
  - APTS
  - PhD
output:
  blogdown::html_page:
    toc: true
    number_sections: false
    toc_depth: 1
---

## The assessment

In my previous [post](https://andrmenezes.github.io/post/2023-01-10-apts-university-of-warwick/), I shared my reflections on the first week of APTS at the
University of Warwick. Now, I will go through my solution for the
Statistical Computing assessment.
The assignment, which can be found [here](https://warwick.ac.uk/fac/sci/statistics/apts/students/resources/starcomp-assessment2223.pdf), involves the implementation of an
`R` function for efficiently and stably fitting a generalized linear model (GLM)
for the one-parameter family distribution.

The density (or mass function) of a one-parameter exponential family model
with canonical parameter $\theta$ is of the form
$$
f(y \mid \theta) = \exp\left\{\theta y - b(\theta) + c(y)\right\}
$$
where $b(\cdot)$ and $c(\cdot)$ are known scalar functions.


The linear predictor $\eta = \mathbf{X}\boldsymbol{\beta}$ in a GLM is related
to the model's canonical parameter $\theta$ via a suitable link function.
Moreover, $\mathbf{X}$ is an $n \times p$ design matrix with known values,
$\boldsymbol{\beta}$ is a $p$-dimensional set of regression parameters, and
$\mathbf{y} = (y_1, \ldots, y_n)$ is an $n$-dimensional vector of observations.

Below, I will outline the main ideas of the solution and compare the estimated
coefficients with those from the R's `glm` function.
You can access a file with the detailed solution
[here](/pdf/solution_statistical_computing_assessment.pdf). 

## The solution

In the first two items, we should write the log-likelihood in vector form and
derive the first and second derivatives with respect to $\boldsymbol{\beta}$.
I will skip the technical details, although they can be found in the detailed
solution [file](/pdf/solution_statistical_computing_assessment.pdf).
The log-likelihood, score vector, and hessian matrix are respectively given by

\begin{eqnarray}
\ell(\boldsymbol{\beta}) &=& \mathbf{y}^\top\,\mathbf{X}\boldsymbol{\beta} - \mathbf{1}^\top\,b(\mathbf{X}\boldsymbol{\beta}) \\
\Delta \ell(\boldsymbol{\beta}) &=& \mathbf{X}^\top\left[\mathbf{y} - b^\prime(\mathbf{X}\boldsymbol{\beta})\right] \\
\Delta^2\ell(\boldsymbol{\beta}) &=& -\mathbf{X}^\top\,\mathrm{diag}(\boldsymbol{\omega})\,\mathbf{X}
\end{eqnarray}
where $b(\cdot)$ and its derivatives are vectorized function and
$\boldsymbol{\omega} = b^{\prime\prime}(\mathbf{X}\boldsymbol{\beta})$.

The score vector and hessian matrix are required in order to apply the
Newton-Rapshon (NR) algorithm to fit the the GLM model.
Given a initial guess,
$\boldsymbol{\beta}^{(0)} = (\beta_0^{(0)}, \beta_1^{(0)}, \ldots, \beta_{p-1}^{(0)})$,
the NR update schema for the one-parameter exponential family model can be
written as follows:


\begin{align*}
\boldsymbol{\beta}^{(k+1)} &= \boldsymbol{\beta}^{(k)} + \left[\mathbf{X}^\top\,\mathrm{diag}(\boldsymbol{\omega}_k)\,\mathbf{X}\right]^{-1}\,\mathbf{X}^\top\,\left[\mathbf{y} - b^\prime(\mathbf{X}\,\boldsymbol{\beta}_k)\right] \\
&= \boldsymbol{\beta}^{(k)} + \left[\mathbf{X}^\top\,\mathrm{diag}(\mathbf{w}_k)\,\mathbf{X}\right]^{-1}\,\mathbf{X}^\top\,\mathbf{z}_k
\end{align*}
where
$\mathbf{w}_k = b^{\prime\prime}(\mathbf{X}\,\boldsymbol{\beta}_k)$ 
and
$\mathbf{z}_k = \mathbf{y} - b^\prime(\mathbf{X}\,\boldsymbol{\beta}_k)$.

By defining $\mathbf{X}_k = \textrm{diag}\left\{\mathbf{w}_k\right\}^{1/2}\,\mathbf{X}$,
we can rewrite the NR update in a more efficient computational form using the
QR decomposition of $\mathbf{X}_k = \mathbf{Q}_k\,\mathbf{R}_k$, where
$\mathbf{R}_k$ is an upper triangular $n\times p$ matrix and $\mathbf{Q}_k$
has the same dimension of $\mathbf{X}_k$ with orthogonal columns such that
$\mathbf{Q}_k^\top\,\mathbf{Q}_k = \mathbf{I}_{n\times p}$.
Also, note that $\textrm{diag}\left\{\mathbf{w}_k\right\} = \textrm{diag}\left\{\mathbf{w}_k\right\}^{1/2}\,\textrm{diag}\left\{\mathbf{w}_k\right\}^{1/2}$,
then the NR update can be set up as:
\begin{align*}
\boldsymbol{\beta}^{(k+1)}
&= \boldsymbol{\beta}^{(k)} + \left[\mathbf{X}^\top\,\mathrm{diag}(\mathbf{w}_k)^{1/2}\,\mathrm{diag}(\mathbf{w}_k)^{1/2}\,\mathbf{X}\right]^{-1}\,\mathbf{X}^\top\,\mathbf{z}_k \\
&= \boldsymbol{\beta}^{(k)} + \left[\mathbf{X}^\top\,\mathrm{diag}(\mathbf{w}_k)^{1/2}\,\mathbf{X}_k\right]^{-1}\,\mathbf{X}^\top\,\mathbf{z}_k \\
&= \boldsymbol{\beta}^{(k)} + \left[\mathbf{X}^\top\,\mathrm{diag}(\mathbf{w}_k)^{1/2}\,\mathbf{Q}_k\,\mathbf{R}_k\right]^{-1}\,\mathbf{X}^\top\,\mathbf{z}_k \\
&= \boldsymbol{\beta}^{(k)} + \mathbf{R}_k^{-1}\,\mathbf{Q}_k^{-1}\,\left[\mathrm{diag}(\mathbf{w}_k)\right]^{-1/2}\left(\mathbf{X}^\top\right)^{-1}\,\mathbf{X}^\top\,\mathbf{z}_k \\
&= \boldsymbol{\beta}^{(k)} + \mathbf{R}_k^{-1}\,\mathbf{Q}_k^{\top}\,\left[\mathrm{diag}(\mathbf{w}_k)\right]^{-1/2}\,\mathbf{z}_k.
\end{align*}

The matrix multiplication of $\mathrm{diag}(\mathbf{w}_k)$ and $\mathbf{z}_k$
can be expressed as a Hadamard (elementwise) product, which is more
computationally efficient. It turns out that
\begin{align*}
\boldsymbol{\beta}^{(k+1)}
&= \boldsymbol{\beta}^{(k)} + \mathbf{R}_k^{-1}\,\mathbf{Q}_k^{\top}\,\left[\mathbf{w}_k^{-1/2}\circ \mathbf{z}_k\right] \\
\boldsymbol{\beta}^{(k+1)} - \boldsymbol{\beta}^{(k)} &= \mathbf{R}_k^{-1}\,\mathbf{Q}_k^{\top}\,\left[\mathbf{w}_k^{-1/2}\circ \mathbf{z}_k\right] \\
\mathbf{R}_k\,(\boldsymbol{\beta}^{(k+1)} - \boldsymbol{\beta}^{(k)}) &= \mathbf{Q}_k^{\top}\,\left[\mathbf{w}_k^{-1/2}\circ \mathbf{z}_k\right].
\end{align*}

#### Computational implementation

We have now implemented a function, `glm1`, which fits a GLM for the one-parameter
family distributions, avoiding the need to perform matrix inversion and create
$n\times n$ matrices of the form $\mathbf{X}^\top \mathbf{X}$.

```{r}
glm1 <- function(y, X, bp, bpp) {
  
  p <- NCOL(X)
  beta_k <- rep(0, p)
  list_beta <- list()
  stop_error <- 1e-6
  j <- 1L
  current_error <- 1
  
  while (current_error > stop_error) {
    eta_k <- X %*% beta_k
    z_k <- y - bp(eta_k)
    w_k <- bpp(eta_k)
    X_k <- drop(w_k^(1/2)) * X
    wz_k <- w_k^(-1/2) * z_k
    qr_out <- qr(X_k)
    Q_k <- qr.Q(qr_out)
    R_k <- qr.R(qr_out)
    a_k <- backsolve(R_k, crossprod(Q_k, wz_k))
    
    # Update the parameter
    list_beta[[j]] <- a_k + beta_k
    current_error <- max(abs(beta_k - list_beta[[j]]))
    beta_k <- list_beta[[j]]
    j <- j + 1L
  }
  
  do.call(cbind, list_beta)
}
```


Because the left hand side of the NR update is upper triangular due to the
$\mathbf{R}_k$ matrix, the `backsolve` function can be used to efficiently
solve a triangular system of linear equations.

The termination criterion is defined as
$\max\left| \boldsymbol{\beta}^{(k)} - \boldsymbol{\beta}^{(k + 1)} \right| < \epsilon$,
where $\epsilon = 10^{-6}$ and the function returns a `data.frame` with the
estimates parameter for each iteration until this criterion is achieved.

#### Cross-checks

I will cross-check my implementation with the R's built `glm` function for the
Bernoulli GLM, which the probability mass function is given by
$$
f(y \mid \theta) = \exp\left\{y\,\theta - \log\left(1 + e^\theta \right)\right\}
$$
where $\theta = \log \left(\dfrac{p}{1 - p}\right)$ and
$b(\theta) = \log\left(1 + e^\theta\right)$. Also,
$b^\prime(\theta) = \dfrac{1}{1 + e^{-\theta}}$ and
$b^{\prime\prime}(\theta) =\dfrac{e^{-\theta}}{(1 + e^{-\theta})^2}$.

A user friendly function, `logReg`, which call the `glm1` is defined
to fit a logistic regression model in a quite similar way as `glm` function. 
```{r}
logReg <- function(formula, data) {
  mf <- model.frame(formula, data = data)
  y <- model.response(mf)
  if (is.factor(y))
    y <- as.numeric(y) - 1
  X <- model.matrix(formula, mf)
  bp <- function(theta) 1 / (1 + exp(-theta))
  bpp <- function(theta) {
    e <- exp(-theta)
    e / (1 + e)^2
  }
  glm1(y, X, bp, bpp)
}
```


The first check concerns a simulated data from $500$ samples of logistic regression model with two
covariates, where the true values of the coefficients are
$\boldsymbol{\beta} = (1.0, -0.5, 0.5)$ and the covariates were simulated from
standard Normal and uniform random variables.

```{r sim-data}
set.seed(69)
n <- 500
X <- cbind(1, rnorm(n), runif(n))
betas <- c(1, -0.5, 0.5)
eta <- drop(X %*% betas)
p <- 1 / (1 + exp(-eta))
y <- rbinom(n, size = 1, prob = p)
sim_data <- data.frame(y = y, x1 = X[, 2], x2 = X[, 3])
```

As shown below, the coefficients estimated from `logReg` and `glm` are the same
to seven decimal places.


```{r est-sim-data}
out_logReg <- logReg(formula = y ~ x1 + x2, data = sim_data)
out_glm <- glm(formula = y ~ x1 + x2, data = sim_data, family = "binomial")
cbind(true = betas,
      logReg = out_logReg[, ncol(out_logReg)],
      glm = coef(out_glm))
```

After performing a second cross-check using the `Pima.tr` data from the `MASS`
package, it was found that the estimated values were identical up to nine decimal
places, confirming that the `glm1` function had been successfully implemented.

```{r est-pima-data}
out_logReg <- logReg(type ~ ., data = MASS::Pima.tr)
out_glm <- glm(type ~ ., data = MASS::Pima.tr, family = "binomial")

cbind(logReg = out_logReg[, ncol(out_logReg)],
      glm = coef(out_glm))
```

