---
title: "Expectation and Maximization algorithm: A toy example"
summary: A toy example of the EM algorithm
author: André Felipe Berdusco Menezes
date: '2022-12-01'
draft: false
categories:
  - EM algorithm
  - maximum likelihood
  - latent variables
output:
  blogdown::html_page:
    toc: true
    number_sections: false
    toc_depth: 1
---

## The problem

The Expectation and Maximization (EM) algorithm is a general procedure to find
the maximum likelihood estimates when the phenomena of interest is describe by
a model with missing values or latent variables. 

In this post I will discus the EM algorithm thorough the following example.
Assume that we observe $X_1, \ldots, X_n$ and know that
$$
X_i = I_{\left\{Z_i > r\right\}} =
\begin{cases}
1 & \textrm{if } Z_i > r \\ 
0 & \textrm{if } Z_i \leq r
\end{cases}
$$
where $r$ is known and $Z_i \sim N(\mu, \sigma^2) \quad \forall\, i=1, \ldots, n$
are unknown.

The problem is, how can we estimate the parameters 
$\boldsymbol{\theta} = (\mu, \sigma^2)$ using the observed values
$\boldsymbol{x} = (x_1, \ldots, x_n)$ from the random variables
$\mathbf{X} = (X_1, \ldots, X_n)$?


## Solution

First, note that $X_i \sim \textrm{Bernoulli}(p)$ where the probability of
success is
$$
p = P(X_i = 1) = P(Z_i > r) = 1 - \varPhi(r)
$$
where $\varPhi(\cdot)$ is the c.d.f. of standard Normal distribution defined as
$$
\varPhi(x) = \int_{-\infty}^{x} \varphi(t) \mathrm{d}t 
\quad \text{e} \quad 
\varphi(x) = \dfrac{1}{\sqrt{2 \pi} \sigma} \mathrm{e}^{- \dfrac{(x - \mu)^2}{2\sigma^2}}.
$$

To use the EM algorithm we need to find the _complete_ likelihood, that is the
joint distribution of
$(\mathbf{X}, \mathbf{Z}) \sim f(\boldsymbol{x}, \boldsymbol{z})$.
Using conditional probability arguments we can write the joint distribution 
for a single observation as
$$
f_{X_i, Z_i}(x, z) = f_{Z_i \mid X_i = x}(z \mid x) f_{X_i}(x).
$$

We also know that
$$
f_{X_i}(x) = \left[1 - \varPhi\left(r\right)\right]^{x}
\left[\varPhi\left(r\right) \right]^{1 - x},\, \quad \mathrm{para}\, x = 0, 1.
$$

In addition, an important remark is that $Z_i \mid X_i = x_i$ follows a
Bernoulli distribution where the probability of success is given by a
truncated Normal distribution at $r$. Its p.d.f. can be written as
$$
f_{Z_i \mid X_i = x}(z_i \mid x_i) = \left[\dfrac{\varphi(z_i)}{1 - \varPhi(r)}\right]^{x_i}\,\left[\dfrac{\varphi(z_i)}{\varPhi(r)}\right]^{1 - x_i}, \, \quad x_i = 0, 1.
$$

Hence, the joint distribution of $(X_i, Z_i)$ is given by
$$
f_{X_i, Z_i}(x_i, z_i) = \left[\dfrac{\varphi(z_i)}{1 - \varPhi(r)}\right]^{x_i}\,\left[\dfrac{\varphi(z_i)}{\varPhi(r)}\right]^{1 - x_i}\,
\left[1 - \varPhi\left(r\right)\right]^{x_i}
\left[\varPhi\left(r\right) \right]^{1 - x_i} = \varphi(z_i).
$$

And the complete likelihood for a sample of size $n$ is given by
$$
L_c(\boldsymbol{\theta} \mid \boldsymbol{z}, \boldsymbol{x}) = \prod_{i=1}^n \varphi(z_i) =
\left(2 \pi \sigma^2\right)^{-n/2}\,\exp\left\{-\dfrac{1}{2\sigma^2}\,\sum_{i=1}^n (z_i - \mu)^2\right\}
$$
where $\boldsymbol{\theta} = (\mu, \sigma^2)$

The E step of EM algorithm is to calculate the expectation of the log-likelihood
with respect to the distribution of $(Z_i \mid X_i = x_i)$. Given an initial
value $\boldsymbol{\theta}_0 = (\mu_0, \sigma^2_0)$ we have
$$
\begin{eqnarray}
Q(\boldsymbol{\theta} \mid \boldsymbol{\theta}_0, \boldsymbol{x}) &=& 
-\dfrac{n}{2}\log(2 \pi \sigma^2) - \dfrac{1}{2\,\sigma^2}\sum_{i=1}^n\left[ 
\mathbb{E}(Z_i^2 \mid  x_i, \boldsymbol{\theta}_0) - 2 \mu \mathbb{E}(Z_i \mid x_i, \boldsymbol{\theta}_0) + \mu^2
\right] \nonumber \\ \nonumber
&=&
-\dfrac{n}{2}\log(2 \pi \sigma^2) - \dfrac{1}{2\,\sigma^2}\left[\sum_{i=1}^n 
h_i(\boldsymbol{\theta}_0) - 2 \mu \sum_{i=1}^n k_i(\boldsymbol{\theta}_0) + n\,\mu^2\right].
\end{eqnarray}
$$
where
$k_i(\boldsymbol{\theta}_0) = \mathbb{E}(Z_i \mid x_i, \boldsymbol{\theta}_0)$ 
and
$h_i(\boldsymbol{\theta}_0) =\mathbb{E}(Z_i^2 \mid x_i, \boldsymbol{\theta}_0)$.

Now, in the M step we should maximize
$Q(\boldsymbol{\theta} \mid \boldsymbol{\theta}_0, \boldsymbol{x})$. Note that

$$
\begin{eqnarray}
\dfrac{\partial}{\partial \mu} Q(\boldsymbol{\theta} \mid \boldsymbol{\theta}_0, \boldsymbol{x}) &=& 
-\dfrac{1}{2\,\sigma^2}\left[2 \sum_{i=1}^n k_i(\boldsymbol{\theta}_0) + 2n\mu\right]
\label{eq:mu}  \\ \label{eq:sigma2}
\dfrac{\partial}{\partial \sigma^2} Q(\boldsymbol{\theta} \mid \boldsymbol{\theta}_0, \boldsymbol{x}) &=& 
-\dfrac{n}{2 \sigma^2} + \dfrac{1}{2\,(\sigma^2)^2}\left[\sum_{i=1}^n 
h_i(\boldsymbol{\theta}_0) - 2 \mu \sum_{i=1}^n k_i(\boldsymbol{\theta}_0) + n\,\mu^2\right].
\end{eqnarray}
$$

Setting the above equations to zero and solving for $\mu$ and $\sigma^2$
it turns out that
$$
\begin{eqnarray}
\dfrac{\partial}{\partial \mu} Q(\boldsymbol{\theta} \mid \boldsymbol{\theta}_0, \boldsymbol{x}) = 0
&\rightarrow& \widehat{\mu} = \dfrac{1}{n} \sum_{i=1}^n k_i(\boldsymbol{\theta}_0)
\\
\dfrac{\partial}{\partial \sigma^2} Q(\boldsymbol{\theta} \mid \boldsymbol{\theta}_0, \boldsymbol{x}) = 0
&\rightarrow&
\widehat{\sigma}^2 = \dfrac{1}{n}\left[\sum_{i=1}^n 
h_i(\boldsymbol{\theta}_0) - 2 \widehat{\mu} \sum_{i=1}^n k_i(\boldsymbol{\theta}_0) + n\,\widehat{\mu}^2\right]
\rightarrow
\nonumber \\
&\rightarrow&
\widehat{\sigma}^2 = \dfrac{1}{n}\left[\sum_{i=1}^n 
h_i(\boldsymbol{\theta}_0) - \dfrac{1}{n}\left(\sum_{i=1}^n k_i(\boldsymbol{\theta}_0)\right)^2   \right].
\end{eqnarray}
$$
Now, we should find the expectations, i.e., 
$k_i(\boldsymbol{\theta}_0) = \mathbb{E}(Z_i \mid x_i, \boldsymbol{\theta}_0)$ and 
$h_i(\boldsymbol{\theta}_0) =\mathbb{E}(Z_i^2 \mid x_i, \boldsymbol{\theta}_0)$.
Remembering that $(Z_i \mid X_i = x_i)$ has Bernoulli distribution with the
probability of success being a truncated Normal distribution at $r$.
Thus, for $Z_i \mid X_i = 1$ the moment generating function is

$$
\begin{align*}
M_{Z_i \mid X_i = 1}(t) = \mathbb{E}[\mathrm{e}^{t Z}] & = \dfrac{1}{1-\varPhi(r)} \int_r^{\infty}  \mathrm{e}^{tz}
\phi(z)\mathrm{d}z =
\dfrac{1}{1-\varPhi(r)} \int_r^{\infty}  \mathrm{e}^{tz} 
\dfrac{1}{\sqrt{2 \pi} \sigma} \mathrm{e}^{- \dfrac{(z - \mu)^2}{2\sigma^2}}\mathrm{d}z\\ \\
&=
\exp\left\{\mu t+ \sigma^2 t^2 / 2 \right\}\,\dfrac{1 - \varPhi(r - \sigma t)}{1 - \varPhi(r)}.
\end{align*}
$$
On the other hand, the moment generating function of $Z_i \mid X_i = 1$ is 

$$
\begin{align*}
M(t) = \mathbb{E}[\mathrm{e}^{t Z}] & = \dfrac{1}{\varPhi(r)} \int_{-\infty}^r  \mathrm{e}^{tz}
\phi(z)\mathrm{d}z =
\dfrac{1}{\varPhi(r)} \int_{-\infty}^r  \mathrm{e}^{tz} 
\dfrac{1}{\sqrt{2 \pi} \sigma} \mathrm{e}^{- \dfrac{(z - \mu)^2}{2\sigma^2}}\mathrm{d}z\\ \\
&=
\exp\left\{\mu t+ \sigma^2 t^2 / 2 \right\}\,\dfrac{\varPhi(r - \sigma t)}{\varPhi(r)}.
\end{align*}
$$
Then the first two moments of $Z_i \mid X_i = x_i$ are given by

$$
E(Z_i \mid X_i = x_i) = M^{\prime}(t)|_{t=0} =
\begin{cases}
\mu + \sigma \dfrac{\varphi(r)}{1 - \varPhi(r)}, & \mathrm{if} \, x_i = 1 \\ \\
\mu - \sigma \dfrac{\varphi(r)}{\varPhi(r)}, & \mathrm{if} \, x_i = 0 
\end{cases}
$$
and
$$
E(Z^2_i \mid X_i = x_i) = M^{\prime\prime}(t)|_{t=0} =
\begin{cases}
\mu^2 + \sigma^2 + \sigma(r + \sigma) \dfrac{\varphi(r)}{1 - \varPhi(r)}, & \mathrm{if} \, x_i = 1 \\ \\
\mu^2 + \sigma^2 - \sigma(r + \sigma) \dfrac{\varphi(r)}{\varPhi(r)}, & \mathrm{if} \, x_i = 0.
\end{cases}
$$
Finally, using the expressions of $k_i(\boldsymbol{\theta}_0)$ and $h_i(\boldsymbol{\theta}_0)$
we can implement the EM algorithm to estimate $\boldsymbol{\theta} = (\mu , \sigma^2)$.

In general, the EM steps are the following

1. Given initial guess $\boldsymbol{\theta}_j = (\mu_j, \sigma_j^2)$;
2. Compute:
\begin{eqnarray*}
  \widehat{\mu}_{j+1} &=& \dfrac{1}{n} \sum_{i=1}^n k_i(\boldsymbol{\theta}_{j}) \\ 
  \widehat{\sigma}^2_{j+1} &=& \dfrac{1}{n}\left[\sum_{i=1}^n h_i(\boldsymbol{\theta}_{j}) - \dfrac{1}{n}\left(\sum_{i=1}^n k_i(\boldsymbol{\theta}_j)\right)^2   \right].
\end{eqnarray*}
and then $\boldsymbol{\theta}_{j+1} = (\widehat{\mu}_{j+1}, \widehat{\sigma}^2_{j+1})$.

3. If $\max\left(|\boldsymbol{\theta}_{j} - \boldsymbol{\theta}_{j+1}|\right) < \varepsilon$ then the convergence criteria is satisfied and the EM estimates of
$\boldsymbol{\theta}$ is $\boldsymbol{\theta}_{j+1}$. 

4. Otherwise, update $\boldsymbol{\theta}_{j}$ for $\boldsymbol{\theta}_{j+1}$ and go through to step (ii).

It is noteworthy that this is a naive example where the two steps of EM
algorithm, the expectation and the maximization can be derived analytically.
Generally, this is not the case for more complex problem although it is possible
to obtain some sort of approximations.

## `R` implementation

In the sequel an R implementation is presented. First, we need to define a 
function to compute the first and second moments of the distribution 
$(Z_i \mid X_i = x_i)$. The below `moments` function can do that.

```{r moments}
moments <- function(x, r, theta) {
  mu <- theta[1]
  sigma2 <- theta[2]
  sigma <- sqrt(sigma2)
  if (x == 1) {
    ki <- mu + sigma * dnorm(r, mu, sigma) / 
      pnorm(r, mu, sigma, lower.tail = F)
    hi <- mu^2 + sigma^2 + sigma * (r + mu) * 
      dnorm(r, mu, sigma) / pnorm(r, mu, sigma, lower.tail = F)
  } else {
    ki <- mu - sigma * dnorm(r, mu, sigma) / 
      pnorm(r, mu, sigma)
    hi <- mu^2 + sigma^2 - sigma * (r + mu) * 
      dnorm(r, mu, sigma) / pnorm(r, mu, sigma)
  }
  c(ki, hi)
}
```

Then, the EM algorithm is implemented in the `expectation_maximization` showed
in the sequence. Note that for each iteration we keep the estimates of $\mu$
and $\sigma^2$. The function returns a matrix with the parameter estimates in 
each iteration. The algorithm stops when reach out the convergence criteria,
which is defined as
$$
\max(|\boldsymbol{\theta}_{j} - \boldsymbol{\theta}_{j+1}|) < \epsilon
$$
where $\epsilon = 10^{-5}$.

```{r em}
expectation_maximization <- function(x, r, theta, tol = 1e-5) {
  estimates <- list()
  n <- length(x)
  err <- 1
  j <- 0L
  while (err > tol) {
    E <- sapply(x, moments, r = r, theta = theta)
    s_E <- rowSums(E, na.rm = T)  
    mu_j <- 1 / n * s_E[1]
    sigma2_j <- 1 / n * (s_E[2] - 1 / n * (s_E[1])^2)
    err <- max(abs(theta[1] - mu_j), abs(theta[2] - sigma2_j))
    theta <- c(mu_j, sigma2_j)
    j <- j + 1
    estimates[[j]] <- c(j, theta)
  }
  m <- do.call(rbind, estimates)
  colnames(m) <- c("Iteration", "mu", "sigma2")
  m
}
```

To evaluate the algorithm a simulate example in presented. I simulated
$Z_i \sim N(4, 2)$ and $r = 4$ then the `expectation_maximization` is used
to get the estimates.

```{r an-example}
set.seed(69)
z <- rnorm(100, mean = 4, sd = 2)
r <- 4
x <- ifelse(z > r, 1, 0)
out <- expectation_maximization(x = x, r = r, theta = c(1, 1), tol = 1e-5)
```

The algorithm reach out the convergence with `r max(out[, 1])` iteration,
and the final estimates of $\mu$ and $\sigma^2$ are
`r formatC(out[max(out[, 1]), 2], digits = 3, format = "f")` and
`r formatC(out[max(out[, 1]), 3], digits = 3, format = "f")`, which are close to
the true values $\mu = 4$ and $\sigma^2 = 4$.

The Figure above shows the estimates obtained in each iteration for the
parameters. Interesting to note that the EM estimates for $\mu$ is
underestimate the true value, while for $\sigma^2$ is overestimate.

```{r plotting, fig.align='center', fig.width=12, fig.height=6}
par(mfrow = c(1, 2))
plot(out[, 2], ylim = c(min(out[, 2]), 4), main = expression(mu),
     xlab = "Iteration", ylab = bquote("Estimate of"~mu))
abline(h = 4, col = "red")
plot(out[, 3], main = expression(sigma^2),
     xlab = "Iteration", ylab = bquote("Estimate of"~sigma^2))
abline(h = 4, col = "red")
```

