---
title: Autoregressive time series models
summary: Simulation, estimation and forecast AR(1)
author: André Felipe Berdusco Menezes
date: "2019-12-16"
draft: false
categories:
  - time series
  - autoregressive model
  - Markov chain
  - likelihood
  - statistical inference
output:
  blogdown::html_page:
    toc: true
    number_sections: false
    toc_depth: 1
---

## Introduction

The [autoregressive](https://en.wikipedia.org/wiki/Autoregressive_model)
process are one of the most simple models studied in coursed or time series analysis.
An interessante feature of autoregressive process is their similar with linear regression models.
For instance, the firs order autoregressive, $\mathrm{AR}(1)$, can be definied as follows:
\begin{equation}\label{eq:ar1}
Y_t \sim N(\mu_t, \sigma^2), \quad \text{where} \quad \mu_t = \alpha + \phi\,y_{t-1}
\end{equation}
for $t = 1,\ldots,n$.
Note that $y_{t-1}$ play the role of covariate and the variance is constant along time.

The goal of this post is to show a naive implementation in `R` for simulation, esitmatiopn and forecasting
an $\mathrm{AR}(1)$ model.
Although it is already available thought the functions `arima.sim()` and
`arima()` this exercise provides a understanding of model structure, and
consequently a path to more complex model implementations.

## Simulation
Let us consider the true parameter values as
$\alpha = -2.0$, $\phi = 0.7$, $\sigma^2 = 4.0$ and the sample size
$n = 500$. Since the probability distribution of the $Y_t$ is Normal I will use
`rnorm()` function.

```{r}
set.seed(666)
n <- 500
alpha <- -2.0
sigma <- 2.0
phi <- 0.7
y <- numeric(n)
for(t in 2:n) {
  mu_t <- alpha + phi * y[t-1]
  y[t] <- rnorm(1, mean = mu_t, sd = sigma)
}
ts.plot(y)
```

## Estimation
Given a random sample $\mathbf{y} = (y_1, \ldots, y_n)$ of the model $\mathrm{AR}(1)$,
under normality assumption the likelihood is given by
$$
L(\mathbf{\theta} \mid \mathbf{y}) \propto
(\sigma^2)^{-\frac{n}{2}}\,\exp\left\{-\frac{1}{2\,\sigma^2}\sum_{t=2}^{n}\left[y_t - (\alpha -\phi\,y_{t-1}) \right]^2 \right\}
$$
where $\mathbf{\theta} = (\alpha, \phi, \sigma^2)$.

The maximum likelihood estimates of parameter vector $\mathbf{\theta}$ is obtained
by the maximization of the log-likelihood.

```{r}
log_like <- function(parms, y) {
  n <- length(y)

  phi <- parms[1]
  alpha <- parms[2]
  sigma <- parms[3]

  eta <- numeric(n)

  for(k in 2:n) eta[k] <- phi * y[k-1]

  ll <- sum(dnorm(x = y, mean = eta + alpha, sd = sigma, log = TRUE))
  ll
}
```
For fixed values of $\alpha$ and $\sigma^2$ the profile likelihood of $\phi$
for the simulated data has the follwing behaviour
```{r, echo=FALSE, fig.align='center', fig.height=6, fig.width=8}
p <- seq(-1, 1, l = 50)
f <- sapply(1:50, function(i) log_like(parms = c(p[i], alpha, sigma), y = y))
plot(p, f, type = "b", ylab = "l", xlab = expression(phi))
abline(v = p[which.max(f)], col = "red")
```

The numeric maximization of log-likelihood in `R` is performed by the `optim`
function. I use the quasi-newton [BFGS](https://en.wikipedia.org/wiki/Broyden%E2%80%93Fletcher%E2%80%93Goldfarb%E2%80%93Shanno_algorithm) optimization algorithm (`method = "BFGS"`), I stated to maximize the function
by using `control = list(fnscale = -1)` and requested to compute the numerical
hessian (`hessian = TRUE`).
```{r, warning=FALSE}
fit <- optim(par = c(phi, alpha, sigma), fn = log_like, y = y, method = "BFGS",
             control = list(fnscale = -1), hessian = TRUE)
```

The paramter estimated for $\phi$, $\alpha$, and $\sigma$ are, respectively,
```{r, echo=FALSE}
fit$par
```
Very close to the true values:
```{r, echo=FALSE}
c(phi, alpha, sigma)
```

Note that the eigenvalues of hessian matriz are negatives, then such estimates
are in fact the maximum likelihood estimates.
```{r}
eigen(fit$hessian)$values
```

## Forecast
For the $\mathrm{AR}(1)$ model the forecast value at time $t$ is given by

$$
y_{t} = \hat{\alpha} + \hat{\phi}\,y_{t-1}
$$
where $\hat{\alpha}$ and $\hat{\phi}$ are the maximum likelihood estimates of
$\alpha$ and $\phi$, respectively.

The picture below shows the simulated time series and the forecast values of
$\mathrm{AR}(1)$ model.
```{r, fig.align='center', fig.height=6, fig.width=8}
mles <- fit$par
phi_hat <- mles[1]
alpha_hat <- mles[2]
sigma_hat <- mles[3]
yhat <- numeric(n)
for(t in 2:n) yhat[t] <- phi_hat * y[t-1]
ts.plot(y)
lines(yhat + alpha_hat, type = "l", col = "blue", lwd = 2)
```
